---
title: "Filtering Data"
output: html_document
---

## How do I filter data?

### Filter with conditions {.tabset .tabset-fade}

Suppose we want to filter mtcars for cars with more than 4 **cyl** (cylinders).

#### In Python

Filtering dataframes can be achieved by placing a boolean expression in a subset expression using `[` 
```{python echo = TRUE, eval = FALSE}
mtcars[mtcars.cyl > 4]
``` 

```{python py-filter}
print(mtcars.cyl[mtcars.cyl > 4])
```

#### In R

Try the same expression except using `$` to reference a column and observe what happens:
```{r filter-wrong, exercise = TRUE, exercise.eval = TRUE, exercise.lines = 3}

```

```{r filter-wrong-hint, include = FALSE}
mtcars[mtcars$col_name > 4]
```

The problem above is that R subsets columns by default, so R evaluates the condition it produces the following:
```{r}
mtcars$cyl > 4
```

A logical vector which exceeds the number of columns we have available, which leads us to the error above.

To filter for rows, you have to specify the columns in addition to the rows when filtering: 

```{r echo = TRUE, eval = FALSE}
df[row_index, col_index]
``` 

where a blank index signifies all rows or columns. Go ahead and fix the code and enter the correct expression below:

```{r filter, exercise = TRUE, exercise.lines = 3}

```

```{r filter-hint-1}
"The problematic code just needs one additional character for the fix."
```
```{r filter-hint-2}
mtcars[row_index, ]
```

```{r filter-check, include = FALSE}
grade_result(
  pass_if(~ identical(.result, mtcars[mtcars$cyl > 4, ]), "Remember to watch out for that comma!")
)
```