---
title: "Inspection"
output: html_document
---

```{python include = FALSE, context="data"}
# set a python variable to point to mtcars
mtcars = r.mtcars
```

## How do I inspect data?

Let's go over how to inspect a dataframe such as *mtcars*:

```{r}
head(mtcars) %>% kable
```

### Print the first few rows {.tabset .tabset-fade}

#### In Python

You are used to printing the first few rows using `head()`

```{python echo = TRUE, eval = FALSE}
mtcars.head()
```

```{python py-head}
print(mtcars.head())
```

#### In R

An equivalent function `head()` exists. Try printing 6 rows of **mtcars** with R:

```{r head, exercise = TRUE, exercise.lines = 3}

```

```{r head-hint}
"Python calls a method on the dataframe object, whereas R passes the dataframe object to the function"
```

```{r head-check, include = FALSE}
grade_result(
  pass_if(~ identical(.result, head(mtcars)), "Not bad huh?")
)
```

### Print the last few rows {.tabset .tabset-fade}

#### In Python

You are used to printing the last few rows using `tail()`. 

```{python echo = TRUE, eval = FALSE}
mtcars.tail()
```

```{python py-tail}
print(mtcars.tail())
```

#### In R

An equivalent `tail()` function exists in R. Try printing the last few rows of **mtcars** in R:

```{r tail, exercise=TRUE, exercise.lines = 3, message=FALSE}

```

```{r tail-check, include = FALSE}
grade_result(
  pass_if(~ identical(.result, tail(mtcars)), "Great job!")
)
```

