---
title: "Problem Filtering"
output: html_document
---

## The problem with filtering in base R

Now let's look at filtering with the same `df`:

```{r}
df
```

### The problem with filtering using `[` {.tabset .tabset-fade}

#### In Python

We can safely use `[` to filter `df` for rows where **a** is greater than 1 notice what happens:

```{python bracket_handled-1, echo=TRUE, eval=TRUE}
df[df.a > 1]
```

and also where **b** is greater than 1:

```{python bracket_handled-2, echo=TRUE, eval=TRUE}
df[df.b > 1]
```

#### In R

Try using `[` to filter `df` for rows where **a** is greater than 1 notice what happens:

```{r bracket_bad, exercise=TRUE, exercise.lines = 3}

```

Now, try using `[` to filter `df` for rows where **b** is greater than 1:

```{r bracket_bad-2, exercise=TRUE, exercise.lines = 3}

```

If you don't see it yet, the issue for both of these operations is that they both kept rows with `NA` even though we wouldn't normally see them in Python.

Now you will appreciate what `filter` does, our next tidy verb! The syntax for it is `filter(df, condition)` where the column in the condition does not need to be referenced from the data.

```{r filter_good-example-1, echo=TRUE}
filter(df, a > 1)
```
```{r filter_good-example-2, echo=TRUE}
filter(df, b > 1)
```

Notice how we don't run into any issues now even with `NA` in our dataframe, just like Python. 

Try tinkering with `filter` yourself:

```{r filter_good, exercise=TRUE, exercise.lines = 3}

```


