---
title: "summarise"
output: html_document
---

## `summarise`

TODO 


```{r group_by-mutate-chained-exercise, echo=TRUE, exercise=TRUE, exercise.lines=6}

```

```{r group_by-mutate-chained-exercise-hint-1}
"Use a chain of piped calls using %>%"
```

```{r group_by-mutate-chained-exercise-hint-2}
"Don't forget about that gotcha on `arrange`!"
```

```{r group_by-mutate-chained-exercise-check}
grade_result(
  pass_if(~ identical(
    .,
    by_cyl %>% 
      mutate(rank = min_rank(desc(mpg))) %>% 
      select(rank, everything()) %>% 
      arrange(rank, .by_group = TRUE)
    )
  )
)
```

Once you've done that we can clearly see the rows ordered according to the rank which is in turn based on **mpg** for each **cyl** groups 4, 6, and 8.

For most data operations, we operate on groups of variables, so `group_by` is an important addition to our arsenal! This will now set us up for performing basic statistics on the data on the next topic with a new verb called `summarise`.

#### If you want to learn more:

There are more advanced things you can do with `group_by`. For example, we didn't cover some of the variantes of `group_by` like TODO. Check out the [docs](https://dplyr.tidyverse.org/reference/group_by.html) for detail.

