---
title: "slice"
output: html_document
---

## `tibble`

The tidverse features a new simpler version of the `data.frame` in R called a `tibble`. We will cover some important differences between the base R dataframe and the tibble in this topic. This will also help you understand some behavioral differences for the next few topics on the `dplyr` verbs.

Let's go through various operations on **mtcars**.

### What is a `tibble`?

A `tibble` is basically just a dataframe that is much more simple and avoids some of the trappings we covered in the "Indexing rows and columns" topic such as partial name matching and dropping dimensions.

Let's start by converting our `mtcars` into a tibble.

```{r tibble-declaration, exercise=TRUE, echo=TRUE}
mtcars_tbl <- as_tibble(mtcars)
mtcars_tbl


```
```{r include=FALSE, context="data"}
mtcars_tbl <- as_tibble(mtcars)
```

You might have noticed something different for both outputs compared to Python. Where are the rownames?!

**Base R ⇒ Tidy R GOTCHA!** tibbles do not use `row.names()`, which means converting from a `data.frame` object will silently drop its rownames.

The reason behind this is a deeper principle of 'tidy data' which treats data as consisting of `variables` and `observations`. Variables are what you know as columns, while observations are rows (see Hadley Wickham's [paper](https://vita.had.co.nz/papers/tidy-data.html). Base R creates a special column just for the row indices, which violates the tidy data principle so everything in the `tidyverse` world ignores them.

However, in some cases row names might be important for datasets like `mtcars` where the car make and model could be nice to see. We can preserve them using the `rownames_to_column` function in the `tibble` package before converting the dataframe to a tibble:

```{r tibble, echo=TRUE, exercise=TRUE, exercise.lines = 4}
mtcars_tbl <- as_tibble(rownames_to_column(mtcars))
mtcars_tbl

```

```{r include=FALSE, context="data"}
mtcars_tbl <- as_tibble(rownames_to_column(mtcars))

```

Row names are back!

### Indexing column(s)

Tibbles behave the same as data.frame when using `[[`:

```{r tibble-subsetting-single, exercise=TRUE, exercise.lines = 4}
mtcars[['mpg']]
mtcars_tbl[['mpg']]

```

the behavior is also the same when using `$`:

```{r tibble-subsetting-single-dollar, exercise=TRUE, exercise.lines = 4}
mtcars$mpg
mtcars_tbl$mpg

```

However, tibbles are a lot more strict and do not allow partial column name matching:

```{r tibble-subsetting-single-dollar-nopartial, exercise=TRUE, exercise.lines = 4}
mtcars$cy
mtcars_tbl$cy

```
While the base R on the first line executes successfully, the second line with the tibble version does not printing an error:
```
Unknown or uninitialised column: 'cy'.NULL
```

The tibble behavior is a lot like Python now, which should ease you a bit more. 

There are no differences between dataframes and tibbles when indexing multiple columns:
```{r tibble-subsetting-columns-1, exercise=TRUE, exercise.lines = 4}
mtcars[c('mpg','cyl')]
mtcars_tbl[c('mpg','cyl')]

```
Well, except with the slight incovenience of having to explicitly select for the `rownames` column for a tibble:
```{r tibble-subsetting-columns-2, exercise=TRUE, exercise.lines = 4}
mtcars[c('mpg','cyl')]
mtcars_tbl[c('rowname','mpg','cyl')]

```

### Indexing rows and columns

Another added consistency is that operations on tibbles will always return a tibble.

For example, subsetting using the `data[row_indices, col_indices]` form will no longer drop dimensions. We can explicitly check that the returned object is still a tibble by using the `is_tibble` function:

```{r tibble-subsetting-1, exercise=TRUE, exercise.lines = 4}
mtcars_tbl[, 'mpg']
is_tibble(mtcars_tbl[, 'mpg'])

```

The same holds true when we index a single column and some rows:

```{r tibble-subsetting-2, exercise=TRUE, exercise.lines = 3}
mtcars_tbl[1:2, c('mpg')]
is_tibble(mtcars_tbl[1:2, c('mpg')])

```
We still get back a tibble.

From here on out, we will use the term 'dataframe' because that is much more familiar to a Python programmer than a 'tibble'. But keep in mind that the verbs that we will learn about next will act on dataframes and sometimes convert them to tibbles to keep with the tidy data principle.

#### If you want to learn more:

There are other differences we do not cover for the scope of this document. You can dig into the [vignette](https://tibble.tidyverse.org/articles/tibble.html) for further details.

